Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-08-29 paper/paper.tex
update paper

\subsection{Boolean satisfiability Problem}
The Boolean satisfiability problem is the problem of determining if there exists an assignment that satisfies a given Boolean formula.
One can specify a satisfiable problem in the conjunctive normal form (CNF), i.e.\ a conjunction of clauses (or disjunctions of Boolean literals).
Given the alphabet of Boolean variables $V$, a CNF can be formally defined as
\begin{equation}
    {\rm CNF} = \bigwedge_{k=1}^{M} \bigvee_{i=1}^{|C^{(k)}|} C^{(k)}_i
\end{equation}
where $M$ is the number of clauses, $C^{(k)}$ is the $k$th clause and $C^{(k)}_i \in V \cup \{\neg v | v \in V\}$ is the $i$th literal in it.
The standard tensor network can be used to study the counting version of the satisfiability problem~\cite{Biamonte2015}, while in the following, we will show a generic reduction from the problem of solving a CNF to a tensor network contraction for solving more solution space properties.
We first map a Boolean literal $v\in V$ and its negation $\neg v$ to the same degree of freedom $s_v \in \{0, 1\}$.
$s_v = 0$ stands for variable $v$ having value \texttt{false} while $s_a=1$ stands for having value \texttt{true}.
% For example, a $k$-th clause $\neg a \vee b \vee \neg c$ can be mapped to a tensor labeled by $(s_a, s_b, s_c)$.
% \begin{equation}
% T^{(k)} = \left(\begin{matrix}
% \left(\begin{matrix}
% x^{w_k} & x^{w_k} \\
% x^{w_k} & x^{w_k}
% \end{matrix}\right) \\
% \left(\begin{matrix}
% x^{w_k} & x^{w_k} \\
% 1 & x^{w_k}
% \end{matrix}\right)
% \end{matrix}\right),
% \end{equation}
% where $w_k$ is a weight associated with a clause. There is only one entry $(s_a, s_b, s_c) = (1, 0, 1)$ that makes this clause unsatisfied. The tensor network representation is
The tensor network $\mathcal{N}_{CNF}$ can be formally defined as
\begin{equation}\label{eq:sattensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(v)} | v \in V\} \cup \{T^{(k)} | k=1,\ldots,M\} \cup \{P\},\\
    \Sigma &= \{s_v | v \in V\},\\
    O &= P,\\
    \lambda &= S \mapsto \begin{cases}
                    (s_{l(k,1)}, s_{l(k, 2)}, s_{l(k,3))}), & S = C^{(k)}\\
                    (), & S = P
                    \end{cases},
\end{split}
\end{equation}
where a tensor defined on literal $v \in V$ is
\begin{equation}
    W^{(v)} = \left(\begin{matrix}
    1\\
    1_v
    \end{matrix}\right)
\end{equation}
and a tensor defined on caluse $C^{(k)}$ is
\begin{equation}
T^{(k)}_{s_{N(k, 1)}, s_{N(k, 2)}, \ldots, s_{N(k, d(k))}} = \begin{cases}
    x^{}, &C^{(k)} ~is~ satisfied ~by~ (s_{N(k, 1)}, s_{N(k, 2)}, \ldots, s_{N(k, d(k))})\\
    1, &{\rm otherwise}
\end{cases},
\end{equation}
where $N(k, i)$ is the $i$th boolean variables in $C^{(k)}$ and $d(k)$ is the number of boolean variables in it.
$w_k$ is the weight associated with clause $C^{(k)}$.
\begin{theorem}\label{thm:cnfcomplex}
    If the addition and multiplication of tensor elements can be computed in time $O(1)$,
    The tensor network representation of a CNF defined in \Cref{eq:sattensornetwork} can be contracted in time $cc(\mathcal{N}_{CNF}) = O(M)2^{O({\rm tw}(H))}$, where $M$ is the number of clauses,
    $H$ is a hypergraph constructed by mapping a variables $v\in V$ to a vertex and the $k$th clause $C^{(k)}$ to a hyperedge connecting $\{N(k, i)|i=1,\ldots, d(k)\}$.
\end{theorem}
This can be proved by showing the hypergraph $H$ is exactly the line graph of $\mathcal{N}_{CNF}$.
Let $x^{w_k} = x$ and $1_v = 1$, one can get a polynomial, in which the $k$-th coefficient gives the number of assignments that $k$ clauses are satisfied.

\subsection{Set packing}
Suppose one has a finite set $V$ and a list of subsets of $V$, denoted as $S$. Then, the set packing problem asks if some $k$ subsets in $S$ are pairwise disjoint.
It is the hypergraph generalization of the independent set problem, where a set corresponds to a vertex and an element corresponds to a hyperedge.
The generic tensor network for the set packing problem also has a similar form as that for the independent set problem
\begin{equation}\label{eq:packtensornetwork}
\begin{split}
    \mathcal{T} &= \{B^{(v)} | v\in V\} \cup \{W^{(\sigma)} | \sigma \in S\} \cup \{P\},\\
    \Sigma &= \{s_\sigma | \sigma \in S\},\\
    O &= P,\\
    \lambda &= T \mapsto \begin{cases}
                    (s_\sigma), & T = W^{(\sigma)}\\
                    (s_{N(v, 1)}, \ldots, s_{N(v, d(v))}), & T = B^{(v)}\\
                    (), & T = P
                    \end{cases},
\end{split}
\end{equation}
where for each $v \in V$, we have the constraints over sets containing it $N(v) = \{\sigma| \sigma\in S \wedge v \in \sigma\}$ as
\begin{equation}
    B^{(v)}_{N(v, 1), N(v, 2),\ldots, N(v, d(v))} = \begin{cases}
        1, & s_{N(v, 1)}+ s_{N(v, 2)} + \ldots + s_{N(v, d(v))}\leq 1,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}
and the vertex tensor for each $\sigma \in S$
\begin{equation}
    W^{(\sigma)} = \left(\begin{matrix}
        1\\
        x^{w_\sigma}_{\sigma}
    \end{matrix}\right),
\end{equation}
where $N(v, k)$ is the $k$th element in $N(v)$ and $d(v) = |N(v)|$ is the number of elements in $N(v)$.

\subsection{Set covering}
Suppose one has a finite set $V$ and a list of subsets of $V$, denoted as $S$.
The set covering problem aims to find the minimum number of sets in $S$ that incorporate (cover) all of elements in $V$.
To get the generic tensor network representation, we first map a set $\sigma \in S$ to a Boolean degree of freedom $s_\sigma\in\{0, 1\}$.
Then the tensor network representation for the set packing problem can be expressed as
\begin{equation}\label{eq:covertensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(\sigma)} | \sigma \in S\} \cup \{B^{(v)} | v \in V\} \cup \{P\},\\
    \Sigma &= \{s_\sigma | \sigma \in \mathcal{S}\},\\
    O &= P,\\
    \lambda &= T \mapsto \begin{cases}
                    (s_\sigma), & T = W^{(\sigma)}\\
                    (s_{N(v, 1)},s_{N(v, 2)},\ldots,s_{N(\sigma, d(v))}), & T = B^{(v)}\\
                    (), & T = P
                    \end{cases},
\end{split}
\end{equation}
where for each $s_\sigma$, we define a parameterized rank-one tensor indexed by it as
\begin{equation}
W^{(\sigma)} = \left(\begin{matrix}
    1 \\
    x_\sigma^{w_\sigma}
    \end{matrix}\right)
\end{equation}
where $x_\sigma$ is a generic typed variable associated with $\sigma$ and $w_\sigma$ is a positive integer as the weight.
For each element $v \in V$, we can define a constraint over all $s \in S$ containing this element, i.e. $N(v) = \{s | s \in S \land v\in s\}$, as
\begin{equation}
B^{(v)}_{s_{N(v, 1)}, s_{N(v, 2)}, \ldots, s_{N(v, d(v))}} = \begin{cases}
    0, & s_{N(v, 1)}=s_{N(v, 2)}=\ldots=s_{N(v, d(v))}=0,\\
    1, & \text{otherwise}.
\end{cases}
\end{equation}
where $N(v, k)$ is the $k$th element in $N(v)$ and $d(v) = |N(v)|$ is the number of elements in $N(v)$. 
If a subset of $S$ does not include any sets containing element $v$, then the corresponding entry is zero.

>>>>

\subsection{Boolean satisfiability Problem}
The Boolean satisfiability problem is the problem of determining if there exists an assignment that satisfies a given Boolean formula.
One can specify a satisfiable problem in the conjunctive normal form (CNF), i.e.\ a conjunction of clauses (or disjunctions of Boolean literals).
Given the alphabet of Boolean variables $V$, a CNF can be formally defined as
\begin{equation}
    {\rm CNF} = \bigwedge_{k=1}^{M} \bigvee_{i=1}^{|C^{(k)}|} C^{(k)}_i
\end{equation}
where $M$ is the number of clauses, $C^{(k)}$ is the $k$th clause and $C^{(k)}_i \in V \cup \{\neg v | v \in V\}$ is the $i$th literal in it.
The standard tensor network can be used to study the counting version of the satisfiability problem~\cite{Biamonte2015}, while in the following, we will show a generic reduction from the problem of solving a CNF to a tensor network contraction for solving more solution space properties.
We first map a Boolean literal $v\in V$ and its negation $\neg v$ to the same degree of freedom $s_v \in \{0, 1\}$.
$s_v = 0$ stands for variable $v$ having value \texttt{false} while $s_a=1$ stands for having value \texttt{true}.
% For example, a $k$-th clause $\neg a \vee b \vee \neg c$ can be mapped to a tensor labeled by $(s_a, s_b, s_c)$.
% \begin{equation}
% T^{(k)} = \left(\begin{matrix}
% \left(\begin{matrix}
% x^{w_k} & x^{w_k} \\
% x^{w_k} & x^{w_k}
% \end{matrix}\right) \\
% \left(\begin{matrix}
% x^{w_k} & x^{w_k} \\
% 1 & x^{w_k}
% \end{matrix}\right)
% \end{matrix}\right),
% \end{equation}
% where $w_k$ is a weight associated with a clause. There is only one entry $(s_a, s_b, s_c) = (1, 0, 1)$ that makes this clause unsatisfied. The tensor network representation is
The tensor network $\mathcal{N}_{CNF}$ can be formally defined as
\begin{equation}\label{eq:sattensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(v)} | v \in V\} \cup \{T^{(k)} | k=1,\ldots,M\} \cup \{P\},\\
    \Sigma &= \{s_v | v \in V\},\\
    O &= P,\\
    \lambda &= S \mapsto \begin{cases}
                    (s_{l(k,1)}, s_{l(k, 2)}, s_{l(k,3))}), & S = C^{(k)}\\
                    (), & S = P
                    \end{cases},
\end{split}
\end{equation}
where a tensor defined on literal $v \in V$ is
\begin{equation}
    W^{(v)} = \left(\begin{matrix}
    1\\
    1_v
    \end{matrix}\right)
\end{equation}
and a tensor defined on caluse $C^{(k)}$ is
\begin{equation}
T^{(k)}_{s_{N(k, 1)}, s_{N(k, 2)}, \ldots, s_{N(k, d(k))}} = \begin{cases}
    x^{}, &C^{(k)} ~is~ satisfied ~by~ (s_{N(k, 1)}, s_{N(k, 2)}, \ldots, s_{N(k, d(k))})\\
    1, &{\rm otherwise}
\end{cases},
\end{equation}
where $N(k, i)$ is the $i$th boolean variables in $C^{(k)}$ and $d(k)$ is the number of boolean variables in it.
$w_k$ is the weight associated with clause $C^{(k)}$.
\begin{theorem}\label{thm:cnfcomplex}
    If the addition and multiplication of tensor elements can be computed in time $O(1)$,
    The tensor network representation of a CNF defined in \Cref{eq:sattensornetwork} can be contracted in time $cc(\mathcal{N}_{CNF}) = O(M)2^{O({\rm tw}(H))}$, where $M$ is the number of clauses,
    $H$ is a hypergraph constructed by mapping a variables $v\in V$ to a vertex and the $k$th clause $C^{(k)}$ to a hyperedge connecting $\{N(k, i)|i=1,\ldots, d(k)\}$.
\end{theorem}
This can be proved by showing the hypergraph $H$ is the line graph of $\mathcal{N}_{CNF}$.
Let $x^{w_k} = x$ and $1_v = 1$, one can get a polynomial, in which the $k$-th coefficient gives the number of assignments that $k$ clauses are satisfied.

\subsection{Set packing}
Suppose one has a finite set $V$ and a list of subsets of $V$, denoted as $S$. Then, the set packing problem asks if some $k$ subsets in $S$ are pairwise disjoint.
It is the hypergraph generalization of the independent set problem, where a set corresponds to a vertex and an element corresponds to a hyperedge.
The generic tensor network for the set packing problem $\mathcal{N}_{pack}$ also has a similar form as that for the independent set problem
\begin{equation}\label{eq:packtensornetwork}
\begin{split}
    \mathcal{T} &= \{B^{(v)} | v\in V\} \cup \{W^{(\sigma)} | \sigma \in S\} \cup \{P\},\\
    \Sigma &= \{s_\sigma | \sigma \in S\},\\
    O &= P,\\
    \lambda &= T \mapsto \begin{cases}
                    (s_\sigma), & T = W^{(\sigma)}\\
                    (s_{N(v, 1)}, \ldots, s_{N(v, d(v))}), & T = B^{(v)}\\
                    (), & T = P
                    \end{cases},
\end{split}
\end{equation}
where for each $v \in V$, we have the constraints over sets containing it $N(v) = \{\sigma| \sigma\in S \wedge v \in \sigma\}$ as
\begin{equation}
    B^{(v)}_{N(v, 1), N(v, 2),\ldots, N(v, d(v))} = \begin{cases}
        1, & s_{N(v, 1)}+ s_{N(v, 2)} + \ldots + s_{N(v, d(v))}\leq 1,\\
        0, & \text{otherwise}.
    \end{cases}
\end{equation}
and the vertex tensor for each $\sigma \in S$
\begin{equation}
    W^{(\sigma)} = \left(\begin{matrix}
        1\\
        x^{w_\sigma}_{\sigma}
    \end{matrix}\right),
\end{equation}
where $N(v, k)$ is the $k$th element in $N(v)$ and $d(v) = |N(v)|$ is the number of elements in $N(v)$.

\begin{theorem}\label{thm:packcomplex}
    If the addition and multiplication of tensor elements can be computed in time $O(1)$,
    The tensor network representation of a set packing problem defined in \Cref{eq:packtensornetwork} can be contracted in time $cc(\mathcal{N}_{pack}) = O(|S|)2^{O({\rm tw}(H))}$, where $|S|$ is the number of sets,
    $H$ is a hypergraph constructed by mapping a set $\sigma\in S$ to a vertex and element $v \in V$ to a hyperedge connecting sets containing it $N(v)$.
\end{theorem}
This can be proved by showing the hypergraph $H$ is the line graph of $\mathcal{N}_{cover}$.

\subsection{Set covering}
Suppose one has a finite set $V$ and a list of subsets of $V$, denoted as $S$.
The set covering problem aims to find the minimum number of sets in $S$ that incorporate (cover) all of elements in $V$.
To get the generic tensor network representation, we first map a set $\sigma \in S$ to a Boolean degree of freedom $s_\sigma\in\{0, 1\}$.
Then the tensor network representation for the set covering problem $\mathcal{N}_{pack}$ can be expressed as
\begin{equation}\label{eq:covertensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(\sigma)} | \sigma \in S\} \cup \{B^{(v)} | v \in V\} \cup \{P\},\\
    \Sigma &= \{s_\sigma | \sigma \in \mathcal{S}\},\\
    O &= P,\\
    \lambda &= T \mapsto \begin{cases}
                    (s_\sigma), & T = W^{(\sigma)}\\
                    (s_{N(v, 1)},s_{N(v, 2)},\ldots,s_{N(\sigma, d(v))}), & T = B^{(v)}\\
                    (), & T = P
                    \end{cases},
\end{split}
\end{equation}
where for each $s_\sigma$, we define a parameterized rank-one tensor indexed by it as
\begin{equation}
W^{(\sigma)} = \left(\begin{matrix}
    1 \\
    x_\sigma^{w_\sigma}
    \end{matrix}\right)
\end{equation}
where $x_\sigma$ is a generic typed variable associated with $\sigma$ and $w_\sigma$ is a positive integer as the weight.
For each element $v \in V$, we can define a constraint over all $s \in S$ containing this element, i.e. $N(v) = \{s | s \in S \land v\in s\}$, as
\begin{equation}
B^{(v)}_{s_{N(v, 1)}, s_{N(v, 2)}, \ldots, s_{N(v, d(v))}} = \begin{cases}
    0, & s_{N(v, 1)}=s_{N(v, 2)}=\ldots=s_{N(v, d(v))}=0,\\
    1, & \text{otherwise}.
\end{cases}
\end{equation}
where $N(v, k)$ is the $k$th element in $N(v)$ and $d(v) = |N(v)|$ is the number of elements in $N(v)$. 
If a subset of $S$ does not include any sets containing element $v$, then the corresponding entry is zero.

\begin{theorem}\label{thm:covercomplex}
    If the addition and multiplication of tensor elements can be computed in time $O(1)$,
    The tensor network representation of a set covering problem defined in \Cref{eq:covertensornetwork} can be contracted in time $cc(\mathcal{N}_{cover}) = O(|S|)2^{O({\rm tw}(H))}$, where $|S|$ is the number of sets,
    $H$ is a hypergraph constructed by mapping a set $\sigma\in S$ to a vertex and element $v \in V$ to a hyperedge connecting sets containing it $N(v)$.
\end{theorem}
This can be proved by showing the hypergraph $H$ is the line graph of $\mathcal{N}_{cover}$.
